---
title: "Wrangle and Featuring"
editor_options: 
  chunk_output_type: console
---

# Initialize

```{r}

# Load Libraries
library(rlang)
library(glue)
library(dtplyr)
library(tidyverse)
library(lubridate)
library(scales)
library(nflreadr)

library(nflModeler)

library(broom)
library(corrr)

# Set Options
options(scipen = 999)
options(dplyr.summarise.inform = FALSE)

# Define Globals
mode <- 'train_light'
this_week <- 22
this_season <- 2022
seasons <- case_when(mode == 'serve' ~ (this_season - 2),
                     mode == 'train_light' ~ 2016,
                     mode == 'train_full' ~ 2006):2022

```

# Add this to nflModeler

```{r}

pivot_home_away <- function(game_key_df, apply_conventions = T) {

  # Collect home prefix
  home_data <- game_key_df %>%
    # Set team type
    mutate(team_type = 'home') %>%
    # Remove away variables
    select(-starts_with('away_')) %>%
    relocate(!starts_with('home_'))
  
  if (apply_conventions) {
    home_data <- home_data %>%
      # Replace the home with team
      rename_with( ~ str_replace_all(.x, '^home_', 'team_'),
                   starts_with('home_')) %>%
      # Replace everything else with id
      rename_with( ~ glue("id_{.x}"),!starts_with('team_'))
  } else {
    home_data <- home_data %>%
      rename_with(~ str_remove_all(.x, 'home_'))
  }
  
  # Collect away prefix
  away_data <- game_key_df %>%
    # Set team type
    mutate(team_type = 'away') %>%
    # Remove away variables
    select(-starts_with('home_')) %>%
    relocate(!starts_with('away_')) %>%
    # Remove home prefix
    rename_with(~ str_remove_all(.x, '^away_'))
  
  if (apply_conventions) {
    away_data <- away_data %>%
      # Replace the home with team
      rename_with( ~ str_replace_all(.x, '^away_', 'team_'),
                   starts_with('away_')) %>%
      # Replace everything else with id
      rename_with( ~ glue("id_{.x}"),!starts_with('team_'))
    
  } else {
    away_data <- away_data %>%
      rename_with(~ str_remove_all(.x, 'away_data'))
  }
  
  # Bind rows
  combined_data <- bind_rows(home_data, away_data)
  # Arrange on game_id if exists
  # This is not necessary but probably saves code and time
  if ('id_game_id' %in% colnames(combined_data)) {
    combined_data <- combined_data %>%
      rename(id_game = id_game_id) %>%
      arrange(id_game)
  }
  
  return(combined_data)
  
}


```


## Source Scripts

```{r}
# Functions
list.files('Source', pattern = 'R$') %>%
  paste0('Source/', .) %>%
  map(source)
```

## Load Data

```{r}

data <- get_data(seasons = seasons) %>%
  # Hail Mary Criteria
  filter(
    !(
      half_seconds_remaining <= 10 &
        pass_attempt == 1 &
        pass_length == "deep"
    ),
    # Last Play Nonsense Criteria!(
    half_seconds_remaining <= 10 &
      (lateral_reception == 1 | lateral_rush == 1)
  )
)

game_team_info <-nflreadr::load_schedules()

get_game_team_info <- function() {
  
  # No seasons parameter because some metrics need to reference entire time frame
  # For example, coach and QB starts need to consider entire timeline
  
  data <-  nflreadr::load_schedules() %>%
    # Convert to tibble to prevent unexpected coercion from dbplyr, dtplyr
    as_tibble() %>%
    pivot_home_away(apply_conventions = T)
  
  return(data)
  
  
    # Remove redundant or deprecated variables
    select(-c(old_game_id,
              gsis,
              nfl_detail_id,
              pfr,
              pff,
              espn,
              stadium,
              qb_name)) %>%
    # Reciprocate not obvious team level variables
    mutate(across(c(result, spread_line), ~ if_else(team_type == 'away', .x * -1, .x)))
    
    
  return(data)
  
}


game_team_info <- get_game_team_info()




schedule_data <- get_schedule_data() %>%
  # Only using bare bones
  select(
    id_game = game_id,
    id_posteam = posteam,
    team_spread_line = spread_line,
    team_moneyline = moneyline,
    team_n_coach_starts = n_coach_starts
  )

game_priors <- get_game_priors() %>%
  rename(
    id_game = game_id,
    id_posteam = posteam,
    id_playoff = playoff,
    id_div_game = div_game,
    id_site = site
  )

elo_data <- get_elo() %>%
  # Only using bare bones
  select(id_game = game_id,
         id_posteam = posteam,
         team_elo = elo)

```

# Feature Creation

## Drive Collapsing

### Regular Drive Collapsed Statistics

```{r}

# There are about 10 games where large amounts of drive data is missing
# This looks like an NFL API error unfortunately

drive_collapsed_regular <- data %>%
  lazy_dt() %>%
  # Filtering Distorted Results
  filter(
    # Two point attempts
    two_point_attempt == 0, 
    !drive_end_transition %in% c(      
      'end of game',
      'fumble',
      'end_game',
      'end_half',
      'end of half',
      'unknown',
      'fumble_safety',
      'fumble, safety'
    ),
    # Drive end transition is missing for entire games for some reason
    # I'm removing it as a filter option for now
    # !is.na(drive_end_transition),
    rush == 1 | qb_dropback == 1,
  ) %>%
  # Cleaning
  select(game_id, posteam, defteam, drive, wp, wpa) %>%
  drop_na() %>%
  # Summarizing
  group_by(game_id, posteam, defteam, drive) %>%
  summarize(drive_wpa = sum(last(wp), last(wpa)) - first(wp)) %>%
  ungroup() %>%
  # Renaming
  as_tibble() %>%
  rename(id_game = game_id) %>%
  rename_with(~glue("id_{.x}"), c(posteam, defteam))

```

### Drive Collapsed Master


```{r}
master_drive_collapsed <- drive_collapsed_regular %>%
  left_join(calculate_drive_points_oe(data), by = c('id_game', 'drive'))
```

## Game Collapsing and Sub Models

### Deriving Priors and Posteriors

```{r}

adjust_success <- function(data) {

  data %>%
    mutate(
      weight = case_when(
        # Scramble
        qb_scramble == 1 & fumble_lost != 1 ~ 1 + 1.15,
        # Interception
        interception == 1 ~ 1 + (-.38),
        # Non Sack Fumble
        fumble == 1 & sack == 0 & fumble_lost == 1 ~ 1 + (-.23),
        # Goaline
        yardline_100 < 3 & down < 4 ~ 1+-1.34,
        TRUE ~ 1
      ),
      success = weight * success
    )
  
}

res_table <- data %>%
  rename(id_game = game_id,
         id_posteam = posteam)%>%
  # Filter Garbage Time Now
  # Eventually weighted versions of this should be used
  filter(!(qtr == 4 & (wp > .9 | wp < .1))) %>%
  adjust_success() %>%
  drop_na(id_posteam, success, down, epa) %>%
  group_by(id_game, id_posteam) %>%
  summarize(outcome_sr = mean(success)) %>%
  ungroup() %>%
  left_join(
    master_drive_collapsed %>%
      select(id_game,
             id_posteam,
             adj_drive_points) %>%
      drop_na() %>%
      group_by(id_game, id_posteam) %>%
      summarize(points = sum(adj_drive_points)) %>%
      ungroup() %>%
      group_by(id_game) %>%
      group_modify( ~ {
        p_vector <- .x$points
        .x %>%
          mutate(points_allowed = if_else(row_number() == 1, p_vector[2], p_vector[1]))
      }) %>%
      ungroup() %>%
      mutate(points_result = points - points_allowed),
    by = c('id_game', 'id_posteam')
  ) %>%
  # Renaming
  rename_with(~ glue("team_{.x}"), c(points_result))

# These will ensure they're untouched during mutations
priors_and_posterior <- schedule_data %>%
  # Join Game Priors
  left_join(game_priors, by = c('id_game', 'id_posteam')) %>%
  # Join ELO Data
  left_join(elo_data,
            by = c('id_game', 'id_posteam')) %>%
  # Result
  left_join(res_table,
            by = c('id_game', 'id_posteam'))

```

### Lookups

```{r}

# ELO Added
# These are game level features so keep it separate
lookup_elo_added <-
  read_csv("https://projects.fivethirtyeight.com/nfl-api/nfl_elo.csv") %>%
  calculate_elo_added(.schedule_data = schedule_data) %>%
  # Only using bare bones
  select(
    id_game = game_id,
    id_posteam = posteam,
    team_eloa = eloa,
    qb_eloa
  )

```

### Passing Lookup

```{r}

lookup_passing <- data %>%
  generate_qb_metrics(
    .this_week = this_week,
    .this_season = this_season,
    bypass_warnings = T,
    return_extra_circumstances = F
  ) %>%
  ungroup() %>%
  select(-gsis) %>%
  rename_with( ~ glue("qb_{.x}"), .cols = -c(game_id, posteam)) %>%
  # Mutate Some QB Stats to 0 if NA
  # Rushing EPA included here because it is reasonable to be neutral
  # PACR, CPOE and Dakota will be imputed from ELO
  mutate(across(c(qb_starts, qb_elo, qb_rushing_epa), ~ if_else(is.na(.x),0,.x))) %>%
  rename(id_game = game_id,
         id_posteam = posteam)

```

### NGS Receiving

```{r}

lookup_ngs_receiving <- 
  # Hard code seasons
  load_nextgen_stats(seasons = 2016:this_season, stat_type = 'receiving') %>%
  as_tibble() %>%
  # Cleaning
  filter(week != 0) %>%
  select(
    -c(
      season_type,
      player_position,
      player_first_name,
      player_last_name,
      player_jersey_number,
      player_short_name,
      player_display_name,
      player_gsis_id
    )
  ) %>%
  relocate(season, week, posteam = team_abbr) %>%
    rename_with(~ str_replace_all(.x, 'avg_','')) %>%
    # Fixing team name
  dplyr::mutate(
    posteam = gsub("WSH", "WAS", posteam),
    posteam = gsub("LAR", "LA", posteam),
  ) %>%
  # Aggregating
  group_by(season, week, posteam) %>%
  summarize(# Un-weighted, independent metrics
    across(c(cushion, separation), ~ mean(.x)),
    # Target weighted
    across(
      c(
        intended_air_yards,
        percent_share_of_intended_air_yards,
        catch_percentage
      ),
      ~ weighted.mean(.x, targets)
    ),
    # Reception weighted
    across(
      c(yac, expected_yac, yac_above_expectation),
      ~ weighted.mean(.x, receptions)
    )
    ) %>%
  ungroup() %>%
  rename(id_season = season, id_week = week, id_posteam = posteam)


```

### NGS Rushing[Beta]

```{r}

if(F) {
rb_vector <- load_players() %>%
  as_tibble() %>%
  filter(position == 'RB') %>%
  pull(gsis_id)

lookup_ngs_rushing <-
  # Hard code seasons
  load_nextgen_stats(seasons = 2016:2022, stat_type = 'rushing') %>%
  as_tibble() %>%
  # Cleaning
  filter(week != 0) %>% # These are seasonal aggregates
  # Remove non running backs
  # In the future, I could expand these to jet streams, orbits
  filter(player_gsis_id %in% rb_vector) %>%
  select(
    -c(
      season_type,
      player_position,
      player_first_name,
      player_last_name,
      player_jersey_number,
      player_short_name,
      player_display_name,
      player_gsis_id
    )
  ) %>%
  relocate(season, week, posteam = team_abbr) %>%
  # Fixing team name
  dplyr::mutate(
    posteam = gsub("WSH", "WAS", posteam),
    posteam = gsub("LAR", "LA", posteam),
  ) %>%
  # These are mostly empty from 2016:2017
  # Eventually I'll probably impute these
  # I think some of these are just linear models
  select(
    -c(expected_rush_yards,
      rush_yards_over_expected,
      rush_yards_over_expected_per_att,
      rush_pct_over_expected)
  ) %>%
  # Aggregating
  group_by(season, week, posteam) %>%
  summarize(# Attempt Weighted
    across(
      c(
        efficiency,
        percent_attempts_gte_eight_defenders,
        avg_time_to_los,
        avg_rush_yards
      ),
      ~ weighted.mean(.x, rush_attempts)
    ),
    n = sum(rush_attempts)) %>%
  ungroup() %>%
  rename_with(~ glue("rush_{.x}"), .cols = -c(season, week, posteam))

rm(rb_vector)

}

```

### Offensive Game Collapsing

```{r}

offense_game_collapsed <- data %>%
  rename(id_game = game_id) %>%
  rename_with(~ glue("id_{.x}"), .cols = c(posteam, defteam, season, week)) %>%
  # Drop NA on some column
  drop_na(starts_with('id_'), epa, ep, wp, down, qb_dropback) %>%
  # Base Filters
  filter(
    two_point_attempt == 0, # 2PA
    play_type %in% c('run','pass'), # Only normal plays
    qb_kneel != 1, # QB Kneels
    qb_spike != 1, # QB Spikes
    ) %>%
  # Weighting
  adjust_success() %>%
  ## Regular Game Collapsing
  group_by(id_game, id_posteam, id_defteam, id_season, id_week) %>%
  summarize(
    n = n(),
    pass_success = mean(success[qb_dropback==1]),
            rush_success = mean(success[qb_dropback ==0]),
            epa = mean(epa),
    team_wp = mean(wp),
      sr = mean(success)) %>%
  ungroup() %>%
  # Joining to collapsed master drive
  left_join(master_drive_collapsed %>%
      group_by(id_game, id_posteam) %>%
  summarize(
    drive_wpa = mean(drive_wpa, na.rm = T),
    drive_points_oe = mean(drive_points_oe, na.rm = T)
  ) %>%
  ungroup(),
            by = c('id_game', 'id_posteam')) %>%
  # ELO Added
  left_join(lookup_elo_added, by = c('id_game', 'id_posteam')) %>%
  # Binding to Corrected Result
  left_join(
    res_table %>% select(id_game, id_posteam, team_points_result, points),
    by = c('id_game', 'id_posteam')
  ) %>%
  # Look up on NGS Receiving
  left_join(lookup_ngs_receiving, by = c('id_posteam','id_season','id_week')) %>%
  # Some older data has duplicate games
  group_by(id_game) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  # Reversing unit features
  reverse_feature(c(
    where(is.numeric),
    -n,
    -starts_with('id_'),
    -starts_with('team_')
  )) %>%
  # Calculate over expected
  over_expected(where(is.numeric),
                -n,
                -starts_with('team_'),
                -starts_with('id_'),
                team_stat = F) %>%
  over_expected(starts_with('team_'), team_stat = T)
  
```

## Rolling Averages

```{r}

new_week <- schedule_data %>%
  separate(
    col = id_game,
    into = c('id_season', 'id_week', 'id_posteam_2', 'id_defteam_2'),
    sep = '_',
    remove = F
  )  %>%
  select(-c(ends_with('_2'),-id_defteam_2,-id_posteam_2)) %>%
  filter(id_season == this_season,
         id_week == this_week) %>%
  mutate(id_season = as.numeric(id_season),
         id_week = as.numeric(id_week),
         id_defteam = if_else(id_posteam == id_posteam_2, id_defteam_2, id_posteam_2)) %>%
  select(id_game,
         id_posteam,
         id_defteam,
         id_season,
         id_week,
         team_spread_line,
         team_moneyline,
         team_n_coach_starts)

new_week

```

```{r}

rolling_average <- function(x, window_size) {
  n <- n()
  window_size <- min(window_size, n)
  cum_avg <- cumsum(na.omit(x)) / (1:n)
  roll_avg <- ifelse(
    1:n < window_size,
    cum_avg[1:n],
    zoo::rollmean(
      x,
      window_size,
      align = "right",
      fill = NA,
      na.rm = TRUE
    )
  )
  return(roll_avg)
}

rolled <- offense_game_collapsed %>%
  bind_rows(select(new_week, id_game, id_posteam, id_season, id_week)) %>%
  # Remove clinched teams
  remove_clinched_teams() %>% 
  # Rolling
  group_by(id_posteam) %>%
  arrange(id_season, id_week, .by_group = T) %>%
  
  mutate(across(c(where(is.numeric),-c(starts_with('id_'), n)),
                ~ lag(rolling_average(.x * n, window_size = 15)))) %>%
  ungroup() %>%
  # Removing play counts
  select(-n)

rolled %>%
  # Investigate these
  drop_na() %>%
  group_by(id_game, id_posteam) %>%
  filter(n() == 1) %>%
  ungroup() %>%
  # Joining to Priors
  left_join(
    select(
      priors_and_posterior,
      id_game,
      id_posteam,
      team_spread_line,
      team_moneyline,
      team_elo,
      starts_with('outcome')),
      by = c('id_game', 'id_posteam')
    ) %>% 
  select(-c(
    id_game,
    id_posteam,
    id_defteam,
    id_season,
    id_week
  )) %>%
  relocate(starts_with('outcome')) %>%
  correlate() %>% view()
  summarize(mean(abs(outcome_sr), na.rm = T))
  
```

```{r}

pivoted <- rolled %>%
  # Self Join
  inner_join(
    y =  select(
      .,
      id_game,
      id_posteam = id_defteam,
      ends_with('_allowed'),
      starts_with('team_')
    ) %>%
      rename_with(
        ~ glue("{str_replace_all(.x, '_allowed','')}_opponent"),
        -c(id_game, id_posteam)
      ),
    by = c('id_game', 'id_posteam')
  ) %>%
  # Remove Allowed since it's no longer needed
  select(-contains('allowed')) %>%
  # Joining to final result
  left_join(select(res_table, id_game, id_posteam, starts_with('outcome_')),
            by = c('id_game', 'id_posteam')) %>%
  relocate(id_game, id_posteam, id_defteam, id_season, id_week, starts_with('outcome_'))


```

# Interactions and Processing

## Interactions

```{r}

# This takes a really long time to run
like_opponent_differential <- data %>%
  rename(id_game = game_id) %>%
  rename_with(~ glue("id_{.x}"), .cols = c(posteam, defteam, season, week)) %>%
  bind_rows(new_week) %>%
  # This is reducing the data to get the normal result
  calculate_like_opponent_differential()

# Interacting
interacted <- pivoted %>%
  # Unit Level Interactions
  interact_metrics(c(
    where(is.numeric),-c(
      starts_with('team_'),
      starts_with('id_'),
      starts_with('outcome_'),
      ends_with('_opponent')
    )
  ),
  mode = 'unit')  %>%
  # Team Level Interactions
  interact_metrics(c(starts_with('team_'),-ends_with('_opponent')),
                   mode = 'team') %>%
  # Join to QB Lookup
  left_join(lookup_passing, by = c('id_game', 'id_posteam')) %>%
  # Mutate Some QB Stats to 0 if NA
  # I already did this in the lookup table but not all QBs match
  # Rushing EPA included here because it is reasonable to be neutral
  # PACR, CPOE and Dakota will be imputed from ELO
  mutate(across(c(qb_starts, qb_elo, qb_rushing_epa), ~ if_else(is.na(.x), 0, .x))) %>%
  # Join to like result
  left_join(select(like_opponent_differential,-id_defteam),
            by = c('id_game', 'id_posteam')) %>%
  # Including Eras
  mutate(
    era = case_when(
      id_season %in% c(2006:2007) ~ 1,
      # Initial Inaccuracies
      id_season %in% c(2008:2013) ~ 2,
      # Second Era
      id_season %in% c(2014:2016) ~ 3,
      # Third Era
      id_season %in% c(2016:2018) ~ 4,
      # NGS Data
      TRUE ~ 5
    ),
    # Current Era
    # Forcing era to Factor
    era = as.factor(era),
    # Forcing NAN to NA
    # Intended for NGS data
    across(everything(), ~ if_else(is.nan(.x), NA, .x))
  )

  interacted %>%
    correlate() %>%
    relocate(term, starts_with('outcome_')) %>%
    view()

write(interacted)

```

# Model Serving

```{r}

library(tidymodels)

serve_week <- interacted %>%
  filter(id_season == this_season,
         id_week == this_week) %>%
    mutate(era = as.factor(era))

preds <- bind_cols(
  # RF
  read_rds('rec_boost_tree.rds') %>%
    predict(new_data = new_week),
  # Cubist
  read_rds('rec_linear_reg.rds') %>%
    predict(new_data = new_week),
  # SVM
  read_rds('rec_mlp.rds') %>%
    predict(new_data = new_week)
) %>%
  rename(xgb = 1,
         glmnet = 2,
         mlp = 3) %>%
  bind_cols(new_week) %>%
  rowwise() %>%
  mutate(blend_pred = mean(c_across(c(xgb, glmnet, mlp)))) %>%
  ungroup() %>%
  mutate(across(c(xgb, glmnet, mlp), ~ round(.x, digits  = 1)))

```

```{r}

library(gt)
library(gtExtras)
library(ggrepel)
library(ggthemes)

# Give Predictions

preds %>%
  select(
    team = id_posteam,
    opponent = defteam,
    spread = spread_line,
    moneyline = moneyline,
    elo = elo,
    'Spicy' = rf,
    'Safe' = bart,
    'All Considered' = svm
  ) %>%
  gt() %>%
  # Spanners
  gt::tab_spanner(label = 'Teams', columns = c(team, opponent)) %>%
  gt::tab_spanner(label = 'Priors', columns = c(spread, moneyline, elo)) %>%
  gt::tab_spanner(
    label = 'Model Predictions',
    columns = c(Spicy, Safe, 'All Considered',)
  ) %>%
  # Title
  tab_header(title = 'NFL Game Predictions',
             subtitle = glue("Season: {this_season}, Week: {this_week}")) %>%
  gt_theme_538()

# Power Rankings

rolled %>%
  filter(season == this_season,
         week == this_week)  %>%
  arrange(-pythag_points) %>%
  mutate(rank = row_number()) %>%
  select(team = id_posteam, rank, raw = pythag_points) %>%
  gt() %>%
  # Header
  tab_header(title = 'Power Rankings',
             subtitle = glue("Season: {this_season}, Week: {this_week}")) %>%
  # Theme
  gt_theme_538()
  
rolled %>%
  filter(season == this_season,
         id_posteam %in% c(preds$id_posteam, preds$defteam)) %>%
  select(id_posteam, week, pythag_points) %>%
  group_by(id_posteam) %>%
  mutate(label = if_else(week == max(week), id_posteam, NA_character_)) %>%
  ungroup() %>%
  ggplot(aes(week, pythag_points, color = id_posteam)) +
  geom_line() +
  geom_label_repel(aes(label = label),
                   nudge_x = 1,
                   na.rm = TRUE) +
  labs(title = glue("Rankings by Week for {this_season}")) +
  ggthemes::theme_fivethirtyeight() +
  theme(legend.position = "none")

```












